---
title: "Homework Assignment 4: Prioritizing potential aquaculture"
author: "Caitlin Nordheim-Maestas"
date: "`r Sys.Date()`"
format: html
---

<https://eds-223-geospatial.github.io/assignments/HW4.html>

# 0. Setup

## read in data

```{r}
#| message: false
#| warning: false

library(tidyverse) # data wrangling
library(janitor) # data wrangling
library(sf) # for spatial data
library(tmap) # for pretty maps
library(here) # file pathing
library(viridisLite) # colors
library(patchwork) # combine plots
library(stars) # rasters
library(terra) # rasters

# load in data 

############### SST
# combine SST rasters into a raster stack
# use c() or rast()
# check resolution and extent

# average_annual_sst_2008.tif
# average_annual_sst_2009.tif
# average_annual_sst_2010.tif
# average_annual_sst_2011.tif
# average_annual_sst_2012.tif

############### Bathymetry
# bathymetry raster: depth.tif

############### Exclusive Economic Zones
# shapefile for the West Coast EEZ: wc_regions_clean.shp

```

# 1. Workflow for single species

## 1.2 Process data

### sst

```{r}
# data checks

# local algebra to find the mean 
# lapp() or mean()

# global algebra to sibtract 273.15 to convert from from Kelvin to Celsius

```

### get bathymetry to match the sst raster

```{r}
# SAME CRS: data check: ensure same crs

# SAME EXTENT: crop depth raster to match the extent of the SST raster
# can do by bbox or crop

# SAME RESOLUTION: resample to ensure the resolutions match
# using the nearest neighbor approach
# done in discussion section 4

# FINAL CHECK: check that the depth and SST match in resolution, extent, and coordinate reference system
# hint: can the rasters be stacked?
```

## 1.3 Find suitable locations

Find suitable locations for marine aquaculture, we’ll need to find locations that are suitable in terms of both SST and depth

Remember, oysters can only exist between 11-30 degrees C and -70 through 0 depth. Set suitable values to 1 and unsuitable values to 0

Matrix we will build for temp
-Inf, 11, 0 # we want this to be a zero
11, 30, 1 # we want this to be a one
30, Inf, 0 # we want this to be a zero

Matrix we will build for depth
-Inf, -70, 0 # we want this to be a zero
-70, 0, 1 # we want this to be a one
0, Inf, 0 # we want this to be a zero


```{r}
# SST first!

# lets build a matrix

# now reclassify


# Now Depth!

# lets build a matrix

# now reclassify
```

### find locations that satisfy both SST and depth conditions

Tip: The SST and depth rasters should now identify the suitability of locations as 0 or 1. To find locations that have both suitable temperature and depth, you can use map algebra. One idea is to multiply the values of the cells, using the lapp() function.

Why multiply? anything * 0 = 0 and anything * 1 = itself, so we will end up with only a suitable temp AND depth if it is 1.

We will do a lapp of the reclassified sst and depth

```{r}
# lapp of the reclassified sst and depth multiplying them, call the output suitable_cells
```

## 1.4 Determine the most suitable EEZ

We want to determine the total suitable area within each EEZ in order to rank zones by priority. To do so, we need to find the total area of suitable locations within each EEZ.

### mask the suitable_cells in west coast

```{r}
# select suitable cells within West Coast EEZs, mask eez by suitable cells
```

### find area of grid cells

In discussion today?

Zonal algebra (x, y)

x is raster, y needs to be a terra object (spatVector or a raster)

extract()

count grid cells

cellsize()

zonal_output --> eez

```{r}

```

### find the total suitable area within each EEZ

hint: it might be helpful to rasterize the EEZ data

```{r}

```

## 1.5 ggPlot of one species

# 2. Generalizable workflow

Note: per species info like min_sst, max_sst min_depth and max_depth will be parms

# 3. Output map for single species


